---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy (Resolved by David Kelemen)"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(viridis)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}

expeditions_dataset <- read_csv("https://github.com/rfordatascience/tidytuesday/raw/refs/heads/master/data/2020/2020-09-22/expeditions.csv")

expeditions_cleaned <- expeditions_dataset %>%
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != "Other") %>%
  mutate(peak_name = fct_reorder(peak_name, expedition_id, .fun = length))

ggplot(expeditions_cleaned, aes(x = peak_name, fill = season)) +
  geom_bar(position = "stack") +
  scale_fill_viridis(discrete = TRUE) +
  theme_light() +
  labs(
     title = "The 15 most popular peaks stacked by season of expedition",
     y = "Number of expeditions",
     fill = "season"
  ) +
  theme(
    axis.text.x = element_text(hjust = 1),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  ) +
  coord_flip()

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
library(scales)

phds_dataset <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/master/data/2019/2019-02-19/phd_by_field.csv")

phd_aggregated <- phds_dataset %>%
  group_by(year, broad_field) %>%
  summarize(total_phds = sum(n_phds, na.rm = TRUE), .groups = "drop")

ggplot(phd_aggregated, aes(x = year, y = total_phds, color = broad_field)) +
  geom_line(linewidth = 1.2) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  labs(
    title = "Number of awarded Ph.D.-s in the US by year",
    color = "Broad Field"
  ) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank())

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute_dataset <- read_csv("https://github.com/rfordatascience/tidytuesday/raw/refs/heads/master/data/2019/2019-11-05/commute.csv")

commute_aggregated <- commute_dataset %>%
  group_by(state, mode, state_region, state_abb) %>%
  summarize(total_commutes = sum(n, na.rm = TRUE), .groups = 'drop') %>%
  spread(mode, total_commutes)

ggplot(commute_aggregated, aes(x = Walk, y = Bike, color = state_region, label = state_abb)) + 
  geom_point(size = 2) + 
  scale_x_log10(labels = comma_format()) + 
  scale_y_log10(labels = comma_format()) + 
  theme_light() + 
  labs(
    title = "Title number of people walking vs. biking to work in each USA state",
    x = "Number of ppl walking to work (log N)",
    y = "Number of ppl biking to work (log N)",
    color = "State Region"
  ) +
  geom_text(
    data = filter(commute_aggregated, state_abb %in% c("CA", "NY", "IL", "MA", "FL", "WA", "OH", "NJ", "OR", "AZ", "MN", "MI", "IN", "UT", "IA", "CT", "LA", "KS", "NE", "RI", "ID", "MT", "ME", "AK", "ND", "NH", "VT", "DE", "WY")),
    color = "black"
  )

```